Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2022-03-21 Manuscript/00-main.Rmd
Revise simulaiton results
Revise real data analysis

tar_load(manu_path)
tar_load(manu_tbs)

tar_load(WLM_train_dat)
# tar_load(tbl_one)
tar_load(ECB_bamlasso_cv)
tar_load(ECB_SBGAM_cv)
tar_load(WLM_bamlasso_cv)
tar_load(WLM_SBGAM_cv)
tar_load(tbl_real_time)
tar_load(ECB_plot)
tar_load(WLM_plot)
# tar_load(sim_tim_tab)
```

@@

\clearpage
\begin{figure}[h]
\includegraphics{Figs/ECB_plot}
\end{figure}

>>>>

tar_load(manu_path)
tar_load(manu_tbs)
tar_load(manu_figs)

tar_load(WLM_train_dat)
tar_load(ECB_bamlasso_cv)
tar_load(ECB_SBGAM_cv)
tar_load(WLM_bamlasso_cv)
tar_load(WLM_SBGAM_cv)
tar_load(tbl_real_time)
tar_load(ECB_plot)
tar_load(WLM_plot)
```

@@

\clearpage
\begin{figure}[h] 
\includegraphics{Figs/ECB_plot}
\caption{TODO: change this caption}
\label{fig:ECB_fig}
\end{figure}

### 2022-03-21 Manuscript/04-Real_Data.Rmd
Revise real data analysis

In this section, we apply the proposed models BHAM, fitted with the EM-CD algorithm, to analyze two real-world metabolomics datasets where the outcomes are binary and continuous respectively. We demonstrate the improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021], while being computationally efficient (see Table \ref{tab:time_real_data}). [TODO: add a sentenec how much faster.]
\begin{center}
Table \ref{tab:time_real_data} here
\end{center}

\subsection{Emory Cardiovascular Biobank}
\label{sec:ECB}
We use the proposed models BHAM to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There are initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we choose the the top 200 features \bg{with largest variance}. We use 5-knot spline additive models for binary outcome using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ are calculated. BHAM-CD obtains superior AUC, Brier score, and misclassification error in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:ECB_res}).

\begin{center}
Table \ref{tab:ECB_res} here
\end{center}

\subsection{Weight Loss Maintenance Cohort}
We use the proposed models BHAM to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focus on the analysis of one of the three studies included, weight loss maintenance cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation \bg{were previously} performed and documented \bg{by the study publishing team [@Bihlmeyer2021]}. The outcome of interest are standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there are $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the Gaussian outcome are constructed using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, $R^2$, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ are calculated. BHAM obtains superior $R^2$, MSE, and MAE in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:WLM_res}).
\begin{center}
Table \ref{tab:WLM_res} here
\end{center}

>>>>

In this section, we apply the proposed models BHAM to analyze two published metabolomics datasets where the outcomes are binary and continuous respectively. We demonstrate the improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021], while being computationally efficient (see Table \ref{tab:time_real_data}). \bg{BHAM requires roughly 10\% of the computation time of SB-GAM to fit models.}

\subsection{Emory Cardiovascular Biobank}
\label{sec:ECB}
We use the proposed models BHAM to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There are initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we choose the the top 200 features \bg{with largest variance}. We use 5-knot spline additive models for binary outcome using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassification error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ are calculated. BHAM obtains superior AUC, Brier score, and misclassification error in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:ECB_res}). \bg{We plot the 33 features included in the final BHAM model in Figure \ref{fig:ECB_fig}.}

\subsection{Weight Loss Maintenance Cohort}
We use the proposed models BHAM to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focus on the analysis of one of the three studies included, weight loss maintenance cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metabolites in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation \bg{were previously} performed and documented \bg{by the study publishing team \cite{Bihlmeyer2021}}. The outcome of interest are standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there are $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the Gaussian outcome are constructed using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, $R^2$, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ are calculated. BHAM obtains superior $R^2$, MSE, and MAE in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:WLM_res}).

### 2022-03-21 Manuscript/SS_GAM.tex
Revise real data analysis

\textcolor{red}{\subsubsection{Variable Selection Performance}}
\textcolor{red}{Among the set of simulations where the functional forms of the predictors are non-linear, the proposed model, BHAM, has a consistent performance across different dimension and distribution settings (See Table \ref{sec:gaus_var_select} for Gaussian outcomes, and Support Information Table S3 for binomial outcomes): being conservative. The symptoms of conservative variable selection are high precision and low recall, where high precision means that among all the selected variables, high percentage of them are true signals; low recall means that, the model selected small subset among all the active predictions. In other words, BHAM tends to select a smaller set of variables that are truly effective to the outcome. We want to note, the variable selection performance of BHAM is plummeted and not optimized when $p=200$ due to using a generic sequence of $s_0$ that doesn't contain the optimal value, discovered when analyzing the simulation result. Overall, among all the models examined, SB-GAM have the best performance, both high precision and high recall, and yield a high MCC. The performance of another Bayesian model, spikeSlabGAM deteriorates as sparsity grows, particularly when ($p>50$), or for binomial outcomes. The variable selection performance for linear simulations match with prediction performance: BHAM performs great with the Gaussian outcome, while SB-GAM's performance deteriorates.}

\textcolor{red}{Among the high-dimensional methods of comparison, there are two methods that are capable to achieve bi-level selection, the proposed BHAM and spikeSlabGAM. Among the linear simulations, both methods can accurately select the linear components while having a drastically lowered probability to include the nonlinear component in the model. Specifically, spikeSlabGAM have smaller probability to include non-linear component in the model than BHAM. However, this advantage of spikeSlabGAM over BHAM is less obvious among the mixed function form simulations: spikeSlabGAM performs better than BHAM when selecting components when the function form includes only linear or nonlinear component, e.g. function form for $x_3$ and $x_4$. However, the spikeSlabGAM inclines to complete ignore the variable that have more complex function form, e.g. function form for $x_1$ and $x_2$, while BHAM are more likely to include them in the model. This trade-off is determined by the assumption implicitly reflected via the prior hierarchy. We defer an in-depth discussion to the Section \ref{sec:concl}.}

@@

In this section, we apply the proposed models BHAM, fitted with the
EM-CD algorithm, to analyze two real-world metabolomics datasets where
the outcomes are binary and continuous respectively. We demonstrate the
improved prediction performance compared to the other Bayesian
hierarchical additive model, SB-GAM \citep{Bai2021}, while being
computationally efficient (see Table \ref{tab:time_real_data}). {[}TODO:
add a sentenec how much faster.{]}

\begin{center}
Table \ref{tab:time_real_data} here
\end{center}

@@

We use the proposed models BHAM to analyze a metabolic dataset from a
recently published research \citep{Mehta2020} studying plasma
metabolomic profile on the three-year all-cause mortality among patients
undergoing cardiac catheterization. The dataset is publicly available
via \textit{Dryad} \citep{Mehta2020_data}. It contains in total of 776
subjects from two cohorts. As there is a large number of non-overlapping
features among the two cohorts, we use the cohort with larger sample
size (N=454). There are initially 6796 features in the dataset, which is
too large to be practically meaningful to analyze. Hence, we choose the
the top 200 features \textcolor{red}{with largest variance}. We use
5-knot spline additive models for binary outcome using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassication error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM-CD obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(see Table \ref{tab:ECB_res}).

\begin{center}
Table \ref{tab:ECB_res} here
\end{center}

@@

We use the proposed models BHAM to analyze metabolomics data from a
recently published study \citep{Bihlmeyer2021} on the association
between metabolic biomarkers and weight loss, where the dataset is
publicly available \citep{Bihlmeyer2021_data}. In this analysis, we
primarily focus on the analysis of one of the three studies included,
weight loss maintenance cohort \citep{Svetkey2008}, due to the
drastically different intervention effects. In the dataset, 765
metaboliltes in baseline plasma collected were profiled using liquid
chromatography mass spectrometry. Quality control and natural log
transformation \textcolor{red}{were previously} performed and documented
\textcolor{red}{by the study publishing team [@Bihlmeyer2021]}. The
outcome of interest are standardized percent change in insulin
resistance, and hence modeled using a Gaussian model. After removing
missing datapoints and addressing outliers in the data, there are
\(p\)=237 features remaining in the analysis. 5-Knot spline additive
models for the Gaussian outcome are constructed using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
\(R^2\), mean squared error (MSE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and mean
absolute error (MAE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|\) are calculated.
BHAM obtains superior \(R^2\), MSE, and MAE in the out-of-bag samples
compared to SB-GAM (see Table \ref{tab:WLM_res}).

\begin{center}
Table \ref{tab:WLM_res} here
\end{center}

>>>>

\textcolor{red}{\subsubsection{Variable Selection Performance}}
\textcolor{red}{Among the set of simulations where the functional forms of the predictors are non-linear, the proposed model, BHAM, has a consistent performance across different dimension and distribution settings (See Table \ref{sec:gaus_var_select} for Gaussian outcomes, and Support Information Table S3 for binomial outcomes): being conservative. The symptoms of conservative variable selection are high precision and low recall, where high precision means that among all the selected variables, high percentage of them are true signals; low recall means that, the model selected small subset among all the active predictions. In other words, BHAM tends to select a smaller set of variables that are truly effective to the outcome. We want to note, the variable selection performance of BHAM is plummeted and not optimized when $p=200$. Upon further investigation, we discover it's because the generic sequence of $s_0$ used to tune the model doesn't contain the optimal value. Overall, among all the models examined, SB-GAM has the best performance, both high precision and high recall, and yield a high MCC. The performance of another Bayesian model, spikeSlabGAM deteriorates as the sparsity grows, particularly when ($p>50$), or for binomial outcomes. The variable selection performance for linear simulations match with prediction performance: BHAM performs great among the Gaussian scenarios, while the performance of SB-GAM deteriorates.}

\textcolor{red}{Among the high-dimensional methods of comparison, there are two methods that are capable to achieve bi-level selection, the proposed BHAM and spikeSlabGAM. Among the linear simulations, both methods can accurately select the linear components and have a drastically lowered probability, close to 0, to include the nonlinear component, as anticipated. Specifically, spikeSlabGAM have smaller probability to include non-linear component in the model than BHAM. However, this advantage of spikeSlabGAM over BHAM is less obvious among the nonlinear simulations: spikeSlabGAM performs better than BHAM when selecting components of the function forms that include only linear or nonlinear component, e.g. function forms for $x_3$ and $x_4$. However, spikeSlabGAM inclines to ignore the variable that have more complex function forms, e.g. function forms for $x_1$ and $x_2$. In contrast, BHAM is more likely to include them in the model. This trade-off is determined by the assumption implicitly reflected via the prior hierarchy. We defer an in-depth discussion to the Section \ref{sec:concl}.}

@@

In this section, we apply the proposed models BHAM to analyze two
published metabolomics datasets where the outcomes are binary and
continuous respectively. We demonstrate the improved prediction
performance compared to the other Bayesian hierarchical additive model,
SB-GAM \citep{Bai2021}, while being computationally efficient (see Table
\ref{tab:time_real_data}).
\textcolor{red}{BHAM requires roughly 10\% of the computation time of SB-GAM to fit models.}

@@

We use the proposed models BHAM to analyze a metabolic dataset from a
recently published research \citep{Mehta2020} studying plasma
metabolomic profile on the three-year all-cause mortality among patients
undergoing cardiac catheterization. The dataset is publicly available
via \textit{Dryad} \citep{Mehta2020_data}. It contains in total of 776
subjects from two cohorts. As there is a large number of non-overlapping
features among the two cohorts, we use the cohort with larger sample
size (N=454). There are initially 6796 features in the dataset, which is
too large to be practically meaningful to analyze. Hence, we choose the
the top 200 features \textcolor{red}{with largest variance}. We use
5-knot spline additive models for binary outcome using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassification error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(see Table \ref{tab:ECB_res}).
\textcolor{red}{We plot the 33 features included in the final BHAM model in Figure \ref{fig:ECB_fig}.}

@@

We use the proposed models BHAM to analyze metabolomics data from a
recently published study \citep{Bihlmeyer2021} on the association
between metabolic biomarkers and weight loss, where the dataset is
publicly available \citep{Bihlmeyer2021_data}. In this analysis, we
primarily focus on the analysis of one of the three studies included,
weight loss maintenance cohort \citep{Svetkey2008}, due to the
drastically different intervention effects. In the dataset, 765
metabolites in baseline plasma collected were profiled using liquid
chromatography mass spectrometry. Quality control and natural log
transformation \textcolor{red}{were previously} performed and documented
\textcolor{red}{by the study publishing team \cite{Bihlmeyer2021}}. The
outcome of interest are standardized percent change in insulin
resistance, and hence modeled using a Gaussian model. After removing
missing datapoints and addressing outliers in the data, there are
\(p\)=237 features remaining in the analysis. 5-Knot spline additive
models for the Gaussian outcome are constructed using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
\(R^2\), mean squared error (MSE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and mean
absolute error (MAE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|\) are calculated.
BHAM obtains superior \(R^2\), MSE, and MAE in the out-of-bag samples
compared to SB-GAM (see Table \ref{tab:WLM_res}).
